Scaling up a GF grammar and preparing it for cool stuff for Swedish

### 2011-11-29 doc/report/report.tex
report

\subsection{GF for Swedish}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology, many phrases and clauses. Most of the implementation
is shared with the other scandinavian languages.
Apart form descriping morphology, the resource grammar deals with
word order, agreement, tempus, basic conjunction etc.
Sentence \ref{gfSwe:parseable} produces the parse in figure \ref{gfSwe:parsetree}.
%Covers sentences like
% better sentence where the verb phrase is split up?
\enumsentence{\shortexnt{8}
{Har& han& inte& ätit& de & gula& äpplena& idag}
{He& has& not& eaten& the& yellow& apples& today}} %correct!!!
\label{gfSwe:parseable}
\begin{figure}[h]
\includegraphics[width=60mm]{gfTree.pdf}
\caption{Parse tree for ``Han har inte ätit de gula äpplena idag".}%correct
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit..} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{``han"} is put between the
finite verb and the rest of the phrase. In the code, this is implemented by using a record.
The linearization rule for questions -- clauses with inverted order -- picks the 
finite form: \\
\verb|table { Inv => verb.fin ++ subj ++ neg ++ verb.inf ++ compl ; ... | \\
Relative clauses are also covered: 
\enumsentence{Hon ser pojken som sover\\
              She sees the boy who sleeps}

In addition to the basic resource grammar, additional constructions had been added
to the module \verb|Extra|. The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
The \verb|Extra| module included functions for topologication, although all
of them were not implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
Check wich were here first:
jag ser dig
dig ser jag
dig har jag inte sett
jag har inte sett dig.
how the parse tree is generated, the finite verb, the subject, the rest. usw.
han såg katten han tyckte om
för vem måste jag akta mig
vem måste jag akta mig för

honom ville hon inte tänka på.

@@

Most important now is NP and VP.
Starting from N, get to CN what is it, what can we do now.
To turn in into a NP need to fix determination, by Det. Dets have .. fixed,
but allow variantion in.. 
What is the difference? Some words on differences for pronouns.
Example, use vartiants {}.
There are also Quants, which .... 
Once we have a NP, can use Predets. What they are.

@@

%%PASSIV
Is passive also more used in Swedish?
There are two ways of forming passive verb phrases in Swedish: the 
periphrastic passive, formed by using a modale auxiliary verb 
\enumsentence{Bilden blev tagen av  av rymdsonden Galileo}
\enumsentence{The picture was taken by spacecraft Galileo.}
or by adding an \emph{s} to
the verb:
\enumsentence{Bilden togs av rymdsonden Galileo}
\enumsentence{The picture [took+s] by spacecraft Galileo.}
This second variant, the s-passive, is commonly used in Swedish, some studies
suggest that it is used in more than 80\% of the times \cite{laanemets}.
For the other scandinavian languages, the s-passive can not be used in all tenses for all words.
The periphrastic passive is preferred, especially in spoken langauge.
\enumsentence{Bildet ble tatt av romsonden (NO)}
The resource grammar for Scandinavian therefore allowed/expressed constructions using auxiliary verb. \\
\verb|PassV2 : V2  -> VP | \\
\verb|         äta -> blev äten| \\
This rules allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
turned into comlpete verb phrases; they no longer need an object.
The s-passive has now been added as the standard case for the Swedish GF
grammar. Periphrastics passive is still allowed, but as an alternative rather
than the default.
The grammar also allows using three-place verbs \verb|V3| , as \emph{give} to
form passives:
\enumsentence{Betyget 3 ges till många elever}
\label{ex:passV3}
by letting the first object be `removed' and given by the subject. 
Sentences where the second object is put first in the sentence, like 
\enumsentence{Till många elever ges betyget 3}
is a variant of example \ref{ex:passV3}, \emph{`många elever'} is still the object but put in focus. The grammar
parse this by applying \verb|FocObj|.
%%

%% Clauses
\subsection{The clause-level}
The resource grammar already provided a number of ways to express relative and embedded clauses, such as
\enumsentence{Frågan är vilken färg den har\\
                   Pojken, som är blyg, tystnar\\
                   Han såg kunden som tyckte om sallad\\
                   Jag tänkte på huset vilket hon bodde i} % better example!!

Since those constructions are inspired by other languages, some of them may sound quite
non-natural in Swedish, such as the \verb|RelS|. This combines a sentence \emph{`Hon sover'}
and a relative sentence \emph{`som/vilket är bra'} which wrongly expressed 
\emph{`She sleeps, which is good'} as
\emph{`Hon sover som är bra'}. The error comes from the fact that \emph{`som'} is
embedded in the relative clause. To fix this, an extra parameter was needed to tell whether
\emph{`som'} or \emph{`vilket'} should be used. \\
Another complication stems from \verb|RelCl| which express the english \emph{`such that'}.
The Swedish version \emph{`sådan att'} sounds stelbent, when used outside of logic books.
\enumsentence{Jag vill ha en katt sådan att den inte fäller.} 
An more natural sounding alternative would be
\enumsentence{Jag vill ha en sådan katt som inte fäller.}
and hopefully this will be implemented soon, otherwise: why is it hard and how we can say
'en katt sådan att det regnar'. It is relative but has on empty field, still congruation is
needed.
du är här, vilket jag vet.
i de äpplen/det äpple du äter bor en mask/ som du ser.\\

\subsection{Presentering, Impersonal constantiation(?), Topicalisering}
Formal subjects \cite{SAG-19,H&H-309d} is often used in Swedish
\enumsentence{Det sitter en fågel på taket}
\emph{`There is a bird sitting on the roof'}
or, literally
\emph{`It sits a bird on the roof'}.\\
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.
Transitive verbs may not be used like this
\enumsentence{*Det äter en fågel på taket.}
unless their in passive form
\enumsentence{Det äts en fågel på taket.}
%det finns/saknas/fattas
There are also restrictions on the nounphrase of the subject, it can only
be determined in certain ways
\enumsentence{*Det sitter den fågel på taket. \\
              **Det sitter fåglarna på taket.}
The distinction lays in the definitness of the noun itself, (not the whole noun phrase..)
it must be indefinite, even though determiners are allowed. Jfr
\enumsentence{Det sitter några fåglar på taket\\
              **Det sitter vardera fåglarna på taket}
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
The grammar handles this by looking at the determiner used
\verb|FormalSub : SimpleVP -> Det -> NP -> Cl ;|
The function combines a verb phrase a determiner and a noun phrase in to a clause.
\verb|SimpleVP| is a normal VP, but with the restriction that the main verb must be
intransitiv, a \verb|V|, (and passives?). If the  determiner requires a noun in determinened
form, the clause is put to NONEXIST.

@@

pronouns/Quant. Many, created categories for 'sådan', and 'fler'. Not many in each,
has to decide, one category for each?
Genitive -problem 1 cannot be alone (fixed) 2 some words have id. (no) gen form
           fixing this include restructuring all
Det-utr. not effect for the some Dets, like 'sådan', no difference between utrum and neutrum.
 Can vary in number, since is a NP. Do not want to cut connection to abstract but changing them,
 so added one more. Leads to ambiguities in some cases, but considering that we would want to use
 this for translation, neutr and utr may differ in the other language.

@@

%% Refl 
Could varandra be thought of as reflexive for 'oss'?
varandra and sin/sitt/sig. Not subject, but NP.
Want the following:
\enumsentence{Han såg sina barns skor.}
\enumsentence{Sina vantar hade han glömt på tåget.}
but not
\enumsentence{*Sina vantar var kvar på tåget.}
\enumsentence{Han ger sina pengar till sina syskon.}
\enumsentence{*Jag ger sina pengar till sina syskon.}
\enumsentence{Han är längre än sin kompis.}
\enumsentence{Hon tyckte om alla sina elever.}
\enumsentence{Han såg sin hund och sitt hus}
\label{ex:refGen}
\enumsentence{Han är längre än sin kompis}
\label{ex:ReflAP}
% hon ber sin katt gå - sin subject in clause.

@@

here or in robust: SupCl, det as VP or VV or S for S (remember Foc)

>>>>

\subsection{GF for Swedish}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other scandinavian languages.
Apart form descriping morphology, the resource grammar deals with
word order, agreement, tempus, basic conjunction etc.
\enumsentence{\shortexnt{8}
{Har& han& inte& ätit& de & gula& äpplena& idag?}
{Has& he& not& eaten& the& yellow& apples& today?}} %correct!!!
\label{gfSwe:parseable}
The grammar can construct and parse the sentence in example \ref{gfSwe:parseable}
and the parse tree is represented as in figure \ref{gfSwe:parsetree}.
%Covers sentences like
% better sentence where the verb phrase is split up?
\begin{figure}[h]
\includegraphics[width=60mm]{gfTree.pdf}
\caption{Parse tree for ``Han har inte ätit de gula äpplena idag".}%correct
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit.."} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{``han"} is put between the
finite verb and the rest of the phrase. At the code level, this is implemented
using records.
The linearization rule for questions -- clauses with inverted order -- picks 
the field for the finite verb and appends it to the subject followed by
the negation, infinite part and the complement of the verb phrase: \\
\verb|table { Inv => verb.fin ++ subj ++ verb.neg ++ verb.inf ++ verb.compl ; ... | \\

Relative clauses are also covered: 
\enumsentence{\shortexnt{5}{Hon & ser & pojken & som  &sover}
              {She & sees & the boy & who & sleeps}}
              %{`She sees the sleeping boy'}}

\enumsentence{\shortexnt{6}{Han & ser &katten & han  &tycker & om}
              {He & sees & the cat & he & likes &}}
              %{`She sees the sleeping boy'}}
In addition to the basic resource grammar, additional constructions had been added
to the module \verb|Extra|. The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
%For example, the \verb|Extra| module included
Among those were functions for topologication, although not all
of them were implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
For example, \verb|FocObj| fronted the object as in sentence \ref{gfSwe:apple}.
\enumsentence{\shortex{6}
{Det & äpplet & vill & jag & inte & ha}
{That & apple & want & I & not &}
{I don't want that apple}}\label{gfSwe:apple}
The module provided functions for preposition stranding:
\enumsentence{\begin{tabular}{ll}
Stranded preposition & cf.\\
Vem måste jag akta mig för? & För vem måste jag akta mig?\\
Who do I need to watch out for? & For whom do I need to wath out?\\
\end{tabular}}

%jag ser dig
%dig ser jag
%dig har jag inte sett
%jag har inte sett dig.
%how the parse tree is generated, the finite verb, the subject, the rest. usw.
%han såg katten han tyckte om
%för vem måste jag akta mig
%vem måste jag akta mig för

%honom ville hon inte tänka på.

@@

Most important now is NP and VP.
Starting from N, get to CN what is it, what can we do now.
To turn in into a NP need to fix determination, by Det. Dets have .. fixed,
but allow variantion in.. 
What is the difference? Some words on differences for pronouns.
To form \emph{`fågeln'}, the determiner (DetQuant DefArt NumSg) is used.
Example, use vartiants {}.
There are also Quants, which .... 
Once we have a NP, can use Predets. What they are.

@@

%%PASSIV
\subsubsection{Passive}
There are two ways of forming passive verb phrases in Swedish: the 
\textbf{periphrastic passive}, formed by using a modale auxiliary verb \\
\enumsentence{ \shortexnt{6}
{Bilden &blev& tagen& av & rymdsonden & Galileo}
{The picture & was & taken & by & the spacecraft & Galileo}}
\label{gfPass:peri-pass}
or by adding an \emph{s} to
the verb: \\
\enumsentence{\shortex{5}
{Bilden & togs&  av & rymdsonden&  Galileo}
{The picture&  took+\textbf{s}&  by&the  spacecraft& Galileo}
{``The picture was taken by the spacecraft Galileo"}}
\label{gfPass:s-pass}
This second variant, the \textbf{s-passive}, is commonly used in Swedish, some studies
suggest that it is used in more than 80\% of the times \cite{laanemets}.
It is not as common in the other scandinavian languages however, 
and it can not be used in all tenses for all words. The norweigan 
translation of both sentence \ref{gfPass:peri-pass} and sentence  \ref{gfPass:s-pass} is
%The periphrastic passive is preferred, especially in spoken langauge.
\enumsentence{Bildet ble tatt av romsonden Galileo}
The resource grammar for Scandinavian therefore expressed the passive using auxiliary verb. \\
\verb|PassV2 : V2  -> VP | \\
\verb|         ta -> blev tagen| \\
This rules allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
turned into comlpete verb phrases; they no longer need an object.
The s-passive has now been added as the standard case for the Swedish GF
grammar. Periphrastics passive is still allowed, but as an alternative rather
than the default.
The grammar also allows using three-place verbs \verb|V3| , as \emph{give}, to
form passives:
\enumsentence{\shortex{6}
{Betyget & 3 &ges &till &många &elever}
{Grade & 3 & give+\textbf{s}& to & many & students}
{``Grade 3 is given to many students"}}
\label{ex:passV3}
by letting the first object be `removed' and given by the subject. 
A remark can be made about sentences where the second object is put first in the sentence, like 
\enumsentence{Till många elever ges betyget 3}
This is a variant of example \ref{ex:passV3}, \emph{`många elever'} is still the object but put in focus. The grammar
parse this by applying \verb|FocObj|.
%%

%% Clauses
\subsubsection*{Presentering, Impersonal constructions, Topicalisering}
Formal subjects \cite{SAG-19,H&H-309d} is often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{It & sits & a & bird & on & the roof}
{``There is a bird sitting on the roof''}}
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.
Transitive verbs may not be used like this
\enumsentence{
*Det äter en fågel på taket\\
 ''There is a bird eating on the roof"}

unless their in passive form
\enumsentence{
Det äts en fågel på taket \\
A bird is being eaten on the roof}
A very common example of this is sentences with the verbs \emph{finnas},
\emph{saknas} and \emph{fattas}.
\enumsentence{\shortex{3}
{Det & finns & kaffe}
{It & exist & coffee}
{''There is coffee"}}
\enumsentence{\shortex{3}
{Det & saknas&  kaffe}
{It & misses & coffee}
{''There is no coffee"}}
\enumsentence{\shortex{3}
{Det & fattas & kaffe}
{It & misses & coffee}
{''There is no coffee"}}
A special GF category was needed for this, the \verb|SimpleVP|.

There are also restrictions on the nounphrase of the subject, it can only
be determined in certain ways
\enumsentence{\begin{tabular}{ll}
*Det sitter den fågel på taket. & **Det sitter fåglarna på taket.\\
There are that bird sitting on the roof & There are the birds sitting on the roof \\
\end{tabular}}
The distinction lays in the definitness of the noun: 
it must be indefinite, even though some determiners are allowed.
\enumsentence{Det sitter några fåglar på taket\\
              **Det sitter vardera fåglarna på taket}
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
The GF grammar does not make any distinction between indefinite and definite noun
phrases. Therefore this is handled by looking at the determiner. \\
\verb|FormalSub : SimpleVP -> Det -> NP -> Cl ;|\\
The function combines a verb phrase a determiner and a noun phrase in to a clause.
If the determiner requires a noun in determinened form, the clause is put to NONEXIST.
However, this is not an ideal solution. There are other examples of when we
need to know the definitness of the noun phrase, for example when using
the reflexive possessive pronoun `sin' (see section `Reflexive pronouns' below)
or when using some predeterminers (see section `Quantifiers')

@@

\subsubsection{Quantifiers}
pronouns/Quant. Many, created categories for 'sådan', and 'fler'. Not many in each,
has to decide, one category for each?
Genitive -problem 1 cannot be alone (fixed) 2 some words have id. (no) gen form
           fixing this include restructuring all
Det-utr. not effect for the some Dets, like 'sådan', no difference between utrum and neutrum.
 Can vary in number, since is a NP. Do not want to cut connection to abstract but changing them,
 so added one more. Leads to ambiguities in some cases, but considering that we would want to use
 this for translation, neutr and utr may differ in the other language.

 'de flesta de gula hästarna'

@@

%% Refl 
\subsubsection*{Reflexive pronouns}
Could varandra be thought of as reflexive for 'oss'?
varandra and sin/sitt/sig. Not subject, but NP.
Want the following:
\enumsentence{Han såg sina barns skor.}
\enumsentence{Sina vantar hade han glömt på tåget.}
but not
\enumsentence{*Sina vantar var kvar på tåget.}
\enumsentence{Han ger sina pengar till sina syskon.}
\enumsentence{*Jag ger sina pengar till sina syskon.}
\enumsentence{Han är längre än sin kompis.}
\enumsentence{Hon tyckte om alla sina elever.}
\enumsentence{Han såg sin hund och sitt hus}
\label{ex:refGen}
\enumsentence{Han är längre än sin kompis}
\label{ex:ReflAP}
% hon ber sin katt gå - sin subject in clause.

@@

here or in robust: SupCl, det as VP or VV or S for S (remember Foc)
\subsubsection*{Relative clauses}
The resource grammar already provided a number of ways to express relative and embedded clauses, such as
\enumsentence{Frågan är vilken färg den har\\
                   Pojken, som är blyg, tystnar\\
                   Han såg kunden som tyckte om sallad\\
                   Jag tänkte på huset i vilket hon bodde} % better example!!

Kanske mindre om detta, skriv om vad jag nu la till.
om 'de äpplen'
Since those constructions are inspired by other languages, some of them may sound quite
non-natural in Swedish, such as the \verb|RelS|. This combines a sentence \emph{`Hon sover'}
and a relative sentence \emph{`som/vilket är bra'} which wrongly expressed 
\emph{`She sleeps, which is good'} as
\emph{`Hon sover som är bra'}. The error comes from the fact that \emph{`som'} is
embedded in the relative clause. To fix this, an extra parameter was needed to tell whether
\emph{`som'} or \emph{`vilket'} should be used. \\
Another complication stems from \verb|RelCl| which express the english \emph{`such that'}.
The Swedish version \emph{`sådan att'} sounds stelbent, when used outside of logic books.
\enumsentence{Jag vill ha en katt sådan att den inte fäller.} 
An more natural sounding alternative would be
\enumsentence{Jag vill ha en sådan katt som inte fäller.}
and hopefully this will be implemented soon, otherwise: why is it hard and how we can say
'en katt sådan att det regnar'. It is relative but has on empty field, still congruation is
needed.
du är här, vilket jag vet.
i de äpplen/det äpple du äter bor en mask/ som du ser.\\
